Supplementary material

### 2018-05-30 Manuscript/ms.tex
Equations for gamma EIP and estimating dv0

The differential equations governing the human population are then given as:
\begin{align} 
\frac{dS}{dt} &= bN - bS - \lambda \frac{V_{I}}{N} S + \delta R\\
\frac{dE}{dt} &= \lambda \frac{V_{I}}{N} S - (\rho + b)E\\
\frac{dI}{dt} &= \rho E - (\gamma + b)I\\
\frac{dR}{dt} &= \gamma I - (\delta + b)R
\end{align}
while the equations governing the mosquito (vector) population are:
\begin{align}
\frac{dV_N}{dt} & = r(t) V_N - \phi_q \tau(t) V_N \\
\frac{dV_E}{dt} &= \lambda \frac{I}{N} V_S - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_E\\
\frac{dV_I}{dt} &= \rho_{v}(t) V_E - (d(t) + \phi_q \tau(t)) V_I\\
\frac{dV_C}{dt} & = \phi_q \tau(t) V_N\\
V_S &= V_N - V_E - V_I.
\end{align}

@@

The remaining parameters include the epidemiological parameters ($\rho$, $\gamma$, $\delta$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}, \nu_0, r_0$), the variances of the latent mosquito processes ($\sigma^2_r$, $\sigma^2_{\nu}$), and the remaining measurement parameters ($\phi_q$, $\eta_y$, $\eta_q$).  
Where possible, we specified informative priors on these parameters based on existing laboratory and field studies (see Table 1 for means and Supplemental Material for detailed explanations).

\begin{table}[!ht]
\label{parameters}
\begin{adjustwidth}{-2.25in}{0in} 
\begin{center}
\caption{Model parameters and their values.  Parameters above the rule are fixed, while parameters below the rule are random, with the value giving the prior mean.}
\begin{tabular}{llll}
Parameter & Description & Value & Citation\\
\hline
$N$ & Human population size & 327801 & \cite{vitpop} \\
$1/d$ & Human life-span & 76 years & \cite{vitlong} \\
$\lambda$ & Transmission rate & 4.87 week$^{-1}$ & \cite{Scott2000}\\
$\phi_y$ & Reporting probability & 0.083 & \cite{Silva2016}\\
$1/\rho_{vt}$ & Extrinsic incubation period & $7\exp \left( 0.2 T_t - 8 \right)$ & \cite{Chan2012}\\
$d_0$ & Mean mosquito mortality rate & 1.47 week$^{-1}$ & \cite{Brady2013} \\
$V_{E0}$ & Initial exposed mosquitoes &  0 & \\
$V_{I0}$ & Initial infectious mosquitoes & 0 & \\
\hline
$1/\rho$ & Latent period in host & 0.87 weeks  & \cite{Chan2012}\\
$\gamma$ & Rate of loss of infectiousness & 3.5 week$^{-1}$ & \cite{Nguyet2013}\\
$1/\delta$ & Period of cross-immunity & 97 weeks &  \cite{Reich2013}\\
$\sigma_r$ & Standard deviation of mosquito growth rate forcing & $0^*$ & \\
$\sigma_{\nu}$ & Standard deviation of mosquito mortality rate forcing & $0^*$ & \\
$S_0$ & Proportion initially susceptible & 0.4 & \cite{Cardoso2011a} \\
$E_0$ & Proportion initially exposed & $8\times 10 ^ {-5}$ & \\
$I_0$ & Proportion initially infectious & $8\times 10 ^ {-5}$ & \\
$r_0$ & Initial mosquito population growth rate & 0 & \\
$V_{N0}$ & Initial mosquito population size & $2N$ & \\
$\nu_0$ & Initial unconstrained mosquito mortality rate & 0 & \\
$\phi_q$ & Per-trap mosquito capture rate & $2 \times 10^{-6}$ week$^{-1}$ & \\
$\eta_y$ & Overdispersion of case reports & $0^*$ & \\
$\eta_q$ & Overdispersion of mosquito trap counts & $0^*$ & \\
\end{tabular}
\end{center}
\end{adjustwidth}
$^*$ indicates prior mode, rather than mean.
\end{table}

@@

Given samples from the posterior distribution as obtained above, we simulated the effects of a single pulse of control applied in each week of the first three years of the time series.
Since the city already implements responsive, targeted control with the aim of reducing the disease burden of an existing outbreak, we focused our simulations on exploring the longer-term feedbacks induced by mosquito control and the ability of an intervention to reduce the disease burden in both the current and future years.
To obtain the posterior distribution of the proportion of cases prevented in calendar year $j$ by implementing control in week $i$, for each posterior sample $k$, we simulated the dynamics resulting from reducing the mosquito population by $5\%$ at the beginning of week $i$ (affecting susceptible, exposed, and infectious mosquitoes equally, and leaving the estimated mosquito birth and mortality rates unchanged) .
We then compared the number of cases produced during year $j$ in the control scenario to the same number in the uncontrolled scenario.
A $5\%$ reduction in mosquito abundance was chosen to keep our simulations conservative relative to field estimates of the mortality induced by spraying \cite{Esu2010}, and to avoid pushing the model into the unrealistic range of dengue eradication.

>>>>

The differential equations governing the human population are then given as:
\begin{align} 
\frac{dS}{dt} &= bN - bS - \lambda \frac{V_{I}}{N} S + \delta R\\
\frac{dE}{dt} &= \lambda \frac{V_{I}}{N} S - (\rho + b)E\\
\frac{dI}{dt} &= \rho E - (\gamma + b)I\\
\frac{dR}{dt} &= \gamma I - (\delta + b)R
\end{align}
while the equations governing the mosquito (vector) population are:
\begin{align}
\frac{dV_N}{dt} & = r(t) V_N - \phi_q \tau(t) V_N \\
\frac{dV_{E1}}{dt} &= \lambda \frac{I}{N} V_S - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E1}\\
\frac{dV_{E2}}{dt} &= \rho_{v}(t) V_{E1} - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E2}\\
\frac{dV_{E3}}{dt} &= \rho_{v}(t) V_{E2}  - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E3}\\
\frac{dV_{E4}}{dt} &= \rho_{v}(t) V_{E3}  - (\rho_{v}(t) + d(t) + \phi_q \tau(t))V_{E4}\\
\frac{dV_I}{dt} &= \rho_{v}(t) V_{E4} - (d(t) + \phi_q \tau(t)) V_I\\
\frac{dV_C}{dt} & = \phi_q \tau(t) V_N\\
V_S &= V_N - V_E - V_I.
\end{align}

@@

The remaining parameters include the epidemiological parameters ($\rho$, $\gamma$, $\delta$, $d_0$), the initial conditions of the model ($S_0$, $E_0$, $I_0$, $R_0$, $V_{N0}, \nu_0, r_0$), the variances of the latent mosquito processes ($\sigma^2_r$, $\sigma^2_{\nu}$), and the remaining measurement parameters ($\phi_q$, $\eta_y$, $\eta_q$).  
Where possible, we specified informative priors on these parameters based on existing laboratory and field studies (see Table 1 for means and Supplemental Material for detailed explanations).

\begin{table}[!ht]
\label{parameters}
\begin{adjustwidth}{-2.25in}{0in} 
\begin{center}
\caption{Model parameters and their values.  Parameters above the rule are fixed, while parameters below the rule are random, with the value giving the prior mean.}
\begin{tabular}{llll}
Parameter & Description & Value & Citation\\
\hline
$N$ & Human population size & 327801 & \cite{vitpop} \\
$1/d$ & Human life-span & 76 years & \cite{vitlong} \\
$\lambda$ & Transmission rate & 4.87 week$^{-1}$ & \cite{Scott2000}\\
$\phi_y$ & Reporting probability & 0.083 & \cite{Silva2016}\\
$1/\rho_{vt}$ & Extrinsic incubation period & $7\exp \left( 0.2 T_t - 8 \right)$ & \cite{Chan2012}\\
$V_{E0}$ & Initial exposed mosquitoes &  0 & \\
$V_{I0}$ & Initial infectious mosquitoes & 0 & \\
\hline
$d_0$ & Baseline mosquito mortality rate & 1.47 week$^{-1}$ & \cite{Brady2013} \\
$1/\rho$ & Latent period in host & 0.87 weeks  & \cite{Chan2012}\\
$\gamma$ & Rate of loss of infectiousness & 3.5 week$^{-1}$ & \cite{Nguyet2013}\\
$1/\delta$ & Period of cross-immunity & 97 weeks &  \cite{Reich2013}\\
$\sigma_r$ & Standard deviation of mosquito growth rate forcing & $0^*$ & \\
$\sigma_{\nu}$ & Standard deviation of mosquito mortality rate forcing & $0^*$ & \\
$S_0$ & Proportion initially susceptible & 0.4 & \cite{Cardoso2011a} \\
$E_0$ & Proportion initially exposed & $8\times 10 ^ {-5}$ & \\
$I_0$ & Proportion initially infectious & $8\times 10 ^ {-5}$ & \\
$r_0$ & Initial mosquito population growth rate & 0 & \\
$V_{N0}$ & Initial mosquito population size & $2N$ & \\
$\nu_0$ & Initial unconstrained mosquito mortality rate & 0 & \\
$\phi_q$ & Per-trap mosquito capture rate & $2 \times 10^{-6}$ week$^{-1}$ & \\
$\eta_y$ & Overdispersion of case reports & $0^*$ & \\
$\eta_q$ & Overdispersion of mosquito trap counts & $0^*$ & \\
\end{tabular}
\end{center}
\end{adjustwidth}
$^*$ indicates prior mode, rather than mean.
\end{table}

@@

Given samples from the posterior distribution as obtained above, we simulated the effects of a single pulse of control applied in each week of the first three years of the time series.
Since the city already implements responsive, targeted control with the aim of reducing local mosquito density during an existing outbreak, we focused our simulations on exploring the longer-term feedbacks induced by mosquito control and the ability of an intervention to reduce the disease burden in both the current and future years.
To obtain the posterior distribution of the proportion of cases prevented in calendar year $j$ by implementing control in week $i$, for each posterior sample $k$, we simulated the dynamics resulting from reducing the mosquito population by $5\%$ at the beginning of week $i$ (affecting susceptible, exposed, and infectious mosquitoes equally, and leaving the estimated mosquito birth and mortality rates unchanged) .
We then compared the number of cases produced during year $j$ in the control scenario to the same number in the uncontrolled scenario.
A $5\%$ reduction in mosquito abundance was chosen to keep our simulations conservative relative to field estimates of the mortality induced by spraying \cite{Esu2010}, and to avoid pushing the model into the unrealistic range of dengue eradication.

### 2018-05-30 Manuscript/supplement.tex
Equations for gamma EIP and estimating dv0

@@

\begin{itemize}
\item \textbf{Mean mosquito mortality rate} ($d_0$): Brady \textit{et al.} \cite{Brady2013} estimated \textit{Aedes aegypti} temperature-dependent field survivorship curves from over 190 laboratory and field experiments.
They found that increased mortality in the field generally washes out the effect of temperature, so we use their estimates for 25 degrees C to construct our prior for $d_0$.  
We use nonlinear least squares to fit the exponential cdf with rate $d_0$ to the estimated mortality curve provided by Brady \textit{et al.}, which gives $d_0 = 1.47$.
\\
\item \textbf{Transmission rate} ($\lambda$): The transmission rate only appears multiplied by a term from the mosquito model (either $V_I$ or $V_S$) and is thus effectively non-identifiable when we are also estimating the overall size of the mosquito population.
Because of this, we fix $\lambda$ at a reasonable value from the literature (though our results should not be sensitive to the exact value used).
The transmission rate is the product of the mosquito feeding rate and the probability of transmission given feeding.
Scott \textit{et al.} \cite{Scott2000} estimate that \textit{Aedes aegypti} females take 0.63 blood meals per day in Puerto Rico, and 0.76 blood meals per day in Thailand.
We take the average of these and set the biting rate at 0.69 per day, or 4.87 per week.
There are relatively few reliable estimates of the transmission probability of DENV, but Lambrechts \textit{et al.} \cite{Lambrechts2011} estimate transmission probabilities close to one  for several related flaviviruses at 25 degrees C (roughly the mean temperature for Vitoria).
\\
\item \textbf{Reporting probability} ($\phi_y$): Silva \textit{et al.} \cite{Silva2016} estimate that there are 12 cases of dengue for every case reported to Brazil's Notifiable Diseases Information System (SINAN, the source of our data), so we fix $\phi_y = 1/12$.
\\
\item \textbf{Extrinsic incubation period} ($1/\rho_{vt}$): Temperature is thought to play an important role in the seasonality of dengue, in particular through its influence on the extrinsic incubation period of the virus in the mosquito  (the amount of time until a mosquito becomes infectious after taking an infectious blood meal).
To account for this, we make $1/\rho_v$ a function of temperature, and force it with weekly mean temperature records for Vitoria downloaded from Weather Underground.
We use the temperature relationship from the recent meta-analysis of \cite{Chan2012}, which incorporates a large number of records and accounts for the interval-censored nature of the data usually available from mosquito transmission studies.
To correspond with the assumptions of the differential equation model, we use the relationship obtained under the assumption that the incubation period at any given temperature is exponentially-distributed:
\begin{equation}
\rho_v = 7\exp \left(0.2T - 8\right)\\
\end{equation}
This relationship is similar to those reported in \cite{Focks1995} and \cite{Tjaden2013}, though there is uncertainty about the exact shape of this relationship.
\end{itemize}

>>>>

\item \textbf{Baselin mosquito mortality rate} ($d_0$): Brady \textit{et al.} \cite{Brady2013} estimated \textit{Aedes aegypti} temperature-dependent field survivorship curves from over 190 laboratory and field experiments.
They found that increased mortality in the field generally washes out the effect of temperature, so we use their estimates for 25 degrees C to construct our prior for $d_0$.  
We use nonlinear least squares to fit the exponential cdf with rate $d_0$ to the estimated mortality curve provided by Brady \textit{et al.}, which gives $d_0 = 1.47$.
We specify a gamma prior on the multiplicative scale, with parameters chosen to keep the prior mass with a factor of 2 of the mean:
d_0' \sim \text{Gamma}(10, 10)
where $d_0 = 1.47d_0'$.

@@

\begin{itemize}
\item \textbf{Transmission rate} ($\lambda$): The transmission rate only appears multiplied by a term from the mosquito model (either $V_I$ or $V_S$) and is thus effectively non-identifiable when we are also estimating the overall size of the mosquito population.
Because of this, we fix $\lambda$ at a reasonable value from the literature (though our results should not be sensitive to the exact value used).
The transmission rate is the product of the mosquito feeding rate and the probability of transmission given feeding.
Scott \textit{et al.} \cite{Scott2000} estimate that \textit{Aedes aegypti} females take 0.63 blood meals per day in Puerto Rico, and 0.76 blood meals per day in Thailand.
We take the average of these and set the biting rate at 0.69 per day, or 4.87 per week.
There are relatively few reliable estimates of the transmission probability of DENV, but Lambrechts \textit{et al.} \cite{Lambrechts2011} estimate transmission probabilities close to one  for several related flaviviruses at 25 degrees C (roughly the mean temperature for Vitoria).
\\
\item \textbf{Reporting probability} ($\phi_y$): Silva \textit{et al.} \cite{Silva2016} estimate that there are 12 cases of dengue for every case reported to Brazil's Notifiable Diseases Information System (SINAN, the source of our data), so we fix $\phi_y = 1/12$.
\\
\item \textbf{Extrinsic incubation period} ($1/\rho_{vt}$): Temperature is thought to play an important role in the seasonality of dengue, in particular through its influence on the extrinsic incubation period of the virus in the mosquito  (the amount of time until a mosquito becomes infectious after taking an infectious blood meal).
To account for this, we make $1/\rho_v$ a function of temperature, and force it with weekly mean temperature records for Vitoria downloaded from Weather Underground.
We use the temperature relationship from the recent meta-analysis of \cite{Chan2012}, which incorporates a large number of records and accounts for the interval-censored nature of the data usually available from mosquito transmission studies.
To correspond with the assumptions of the differential equation model, we use the relationship obtained under the assumption that the incubation period at any given temperature is exponentially-distributed:
\begin{equation}
\rho_v = 7\exp \left(0.2T - 8\right)\\
\end{equation}
This relationship is similar to those reported in \cite{Focks1995} and \cite{Tjaden2013}, though there is uncertainty about the exact shape of this relationship.
\end{itemize}
